Ship-based lidar evaluation of Southern Ocean low clouds in the storm-resolving general circulation model ICON and the ERA5 and MERRA-2 reanalyses

### 2025-07-21 manuscript_agu_rev1.tex
Fix grammar

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model, but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction and very low-level ($<$~250~m) cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. In total, about 2300 radiosonde profiles south of 40°S were available. Spatially and temporally colocated profiles were taken from the models. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with ICON, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. Derived thermodynamic [virtual potential temperature ($\theta_v$), lifting condensation level (LCL), etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

@@

An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10~December, 16--17~December 2016; MARCUS (3 days): 8, 10~November, 10~December 2017; MICRE (9~days): 7--8, 29~June, 5, 16~July, 15~August, 17~October 2016, 11~February, 21~March 2017; TAN1502 (1~day): 24~January.

@@

The Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2) is a reanalysis produced by the Global Modeling and Assimilation Office at the NASA Goddard Space Flight Center \cite{gelaro2017}. It uses version 5.12.4 of the Goddard Earth Observing System (GEOS) atmospheric model \cite{rienecker2008,molod2015}. Non-convective clouds (condensation, autoconversion, and evaporation) are parameterized using a prognostic scheme \cite{bacmeister2006}, and sub-grid cloud fraction is determined using total water distribution and a critical RH threshold. The reanalysis output analyzed here is available at a spatial resolution of 0.5° of latitude and 0.625° of longitude, which is about 56 km in the North--South direction and 35 km in the East--West direction at 60°S. The number of vertical model levels is 72. Here, we use the following products: 1-hourly instantaneous 2D single-level diagnostics (M2I1NXASM) for 2-m temperature and humidity; 3-hourly instantaneous 3D assimilated meteorological fields (M2I3NVASM) for cloud quantities, pressure, and temperature; 1-hourly average 2D surface flux diagnostics (M2T1NXFLX) for precipitation; and 1-hourly average 2D radiation diagnostics (M2T1NXRAD) for radiation quantities \cite{merra2}. Vertically resolved fields in M2I3NVASM start at a height of about 60 m, which limits our analysis of fog and very low-level ($<$~250~m) clouds in this reanalysis.

@@

ERA5 \cite{era5} is a reanalysis produced by the ECMWF. It is based on a numerical weather prediction model IFS version CY41R2. It uses the \citeA{tiedtke1993} prognostic cloud scheme and \citeA{forbes2014} for mixed-phase clouds. The horizontal resolution is 0.25° in latitude and longitude, which is about 28~km in the North--South direction and 14~km in the East--West direction at 60°S. Internally, the model uses 137 vertical levels. Here, we use output at 1-hourly instantaneous time intervals, except for radiation quantities, which are accumulations (from these we calculate daily means). Vertically resolved quantities are made available on 37 pressure levels.

@@

In our analysis, we partitioned our dataset by cyclonic activity and stability into multiple subsets to evaluate cloud biases in the context of the main physical controlling processes. The SO is a region of the occurrence of both extratropical and polar cyclones. Cyclonic activity results in cloud formation at the air mass boundaries along the cold and warm fronts, as well as inside the cold sector, after a passing cold sector destabilizes the atmosphere relative to the surface temperature. In the cold front and cold sector, clouds are convectively driven, including deep convection, and the advection of colder air masses over warmer ocean surfaces can trigger convection and subsequent cloud formation. In contrast, warm advection can trigger fog or cloud formation by boundary layer air cooled by the ocean surface until it reaches saturation. More quiescent areas outside of cyclones can also be associated with clouds. These can be, for example, associated with clouds formed by warm or cold advection outside of cyclones, persistent clouds, clouds formed due to diurnal heating or cooling, or due to ocean currents. Boundary layer stability can be expected to be associated with clouds by either allowing convection and turbulence under weak stability, inhibiting convection turbulence under strong stability, and by capping inversion controlling the cloud top height or trapping moist air near the surface and preventing fog dispersion. Therefore, dividing our dataset by these subsets allows us to quantify model biases associated with some of the main physical processes controlling cloud formation, persistence, and dissipation. Other methods of subsetting, such as using the ISCCP pressure--optical thickness diagram \cite{rossow1991,rossow1999,hahn2001} to separate profiles by cloud regimes and other cloud regime classifications \cite{oreopoulos2016,schuddeboom2018}, would be feasible.

We partitioned our data into two mutually exclusive subsets by cyclonic activity. For this purpose, we used a cyclone tracking algorithm to identify extratropical and polar cyclones (ECs and PCs) over the SO in the reanalysis and ICON data. We used the open-source cyclone tracking package CyTRACK \cite{perez-alarcon2024}. Generally, what constitutes an EC is considered relatively arbitrary due to the very large variability of ECs \cite{neu2013}. The CyTRACK algorithm uses mean sea level pressure and wind speed thresholds as well as tracking across time steps to identify cyclone centers and their radii in each time step. With this information, we could classify every location at a given time as either cyclonic or non-cyclonic. Due to a relatively small total area covered by cyclones, as identified by the cyclone center and radius, for every time step and cyclone we defined a cyclonic area as a circle of double the radius identified by CyTRACK centered at the cyclone center. All other areas were defined as non-cyclonic. For identifying cyclones in the observations and the reanalyses, ERA5 pressure and wind fields were used as the input to CyTRACK. This is justified by the fact that the large-scale pressure and wind fields in ERA5 are likely sufficiently close to reality. \citeA{mcerlich2023} have shown that wind is simulated well in ERA5 relative to the WindSat polarimetric microwave radiometer measurements \cite{meissner2009}. For identifying cyclones in ICON, its own pressure and wind fields were used as the input to CyTRACK because ICON is free-running, and thus the pressure and wind fields are different from reality. Subsetting by proximity to cyclones is a relatively crude measure because it does not take into account the different sectors of cyclones, which are commonly associated with different weather situations. However, this was a choice made for simplicity of the analysis, given the quantity of data. \citeA{konstali2024} performed a more complex attribution of precipitation to individual cyclone features.

@@

Fig.~\ref{fig:cyclone-stability}a, b show the geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section~\ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone and is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and ICON agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences, particularly in the highest occurrence rate regions, such as around Cape Adare, which is up to 20\% more cyclonic in ICON, and the Weddell and Bellingshausen Seas, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that ICON is free-running.

@@

When the data are subset by latitude (Fig.~\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone (40--55°S) displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone (between 55°S and the Antarctic coast), and this could be because higher latitudes have a greater prevalence of weakly stable profiles (Fig.~\ref{fig:cyclone-stability}c, d), although more stable profiles populate regions south of 65°S close to the Antarctic coast. Cyclonic activity is also stronger in high-latitude SO, which is typically associated with shallow or deep convection rather than very stable stratification necessary for fog formation. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (very low-level cloud occurrence is still strongly underestimated).

When the data are subset as either cyclonic and non-cyclonic situations (Fig.~\ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a larger amount of observed cloudiness, including peak and total cloud fraction, by about 10\%. In the cyclonic situations, the model vertical profiles of cloud occurrence compare well with observations, but they peak higher by about 200 m and are larger by about 8\%. The reanalyses tend to underestimate cloud occurrence above 1 km by about 5\% and near the surface by about 15\%. Non-cyclonic situations are similar to the general case, also because they form the majority of analyzed profiles (83\%).

When the data are subset by stability (Fig.~\ref{fig:cloud-occurrence}f, g), as defined in Section~\ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly near the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1~km. Physically, conditions of strong stability are associated with the formation of advection fog, such as in situations of warm air advection from the north over a colder sea surface, thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak up to 12\%, underestimated cloud occurrence above 1~km by up to 5\%, and underestimated cloud occurrence near the surface by about 10\% in the reanalyses but not ICON. In situations of weak stability, the bias in ICON is very pronounced, with a much larger peak in cloud occurrence at about 500~m; the reanalyses underestimate cloud occurrence below 1~km, especially near the surface; and MERRA-2 underestimates cloud occurrence more strongly at almost all heights.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/cl_agg_rev1.pdf}
\caption{
Cloud occurrence by height calculated as the average of all voyages and stations and lifting condensation level (LCL) distribution. The LCL is derived from radiosonde profiles and equivalent model profiles, which were not available for all voyages and times. The total cloud fraction (CF), average shortwave (SW), and longwave (LW) and the relative frequency of occurrence (RFO) are shown. The bands are the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentile, calculated from the set of all voyages and stations.
}
\label{fig:cloud-occurrence}
\end{figure}

@@

We also analyzed the daily cloud cover (total cloud fraction) distribution. This is a measure of cloudiness, irrespective of height, calculated over the course of a day (UTC). A cloud detected at any height means that the lidar profile was classified as cloudy; otherwise, it was classified as a clear sky. When all profiles in a day are taken together, the cloud cover for the day is defined as the fraction of cloudy profiles in the total number of profiles. It is expressed in oktas (multiples of 1/8), reflecting the 3-hourly model output of MERRA-2 and ICON, i.e. 8 times per day. The same calculation is done for the lidar observations as for the simulated lidar profiles. We use the term ``okta'' independently of its use in instantaneous synoptic observations, and here it simply means 1/8 (0.125) of the daily cloud cover.

@@

Fig.~\ref{fig:stats-hist-si} shows liquid water path (LWP), and ice water path (IWP) distributions as histograms and their corresponding averages. LWP and IWP are calculated from the mass of water in the column divided by the area of the column, i.e. not just the area of the cloudy portion of the column, as in some definitions. The available observational satellite reference for LWP and IWP over high latitudes is unfortunately very uncertain due to a high solar zenith angle and inability of passive visible and infrared retrievals to detect phase below the cloud top of mixed-phase clouds \cite{huang2006,greenwald2009,seethala2010,eliasson2011,duncan2018,khanal2020}, and this limits our comparison. The LWP distribution shows that all models overestimate cases with near-zero LWP (Fig.~\ref{fig:stats-hist-si}a), which relates to the underestimated total cloud fraction. MERRA-2 shows quite overestimated high-LWP situations, which is most likely related to the too few, too bright problem of simulating lower total cloud fraction but clouds with higher LWP to compensate. IWP (Fig.~\ref{fig:stats-hist-si}b) is somewhat less important radiatively than LWP because of the typically larger and less numerous hydrometeors. Similarly to LWP, the models overestimate situations with near-zero IWP. ERA5 is otherwise simulating the IWP distribution well, but ICON and MERRA-2 underestimate IWP. In the cloudy situations (Fig.~\ref{fig:stats-hist-si}c, d), it can be seen more distinctly that MERRA-2 overestimates moderate (0.05--0.15 kg m$^{-2}$) and high LWP (over 0.15 kg m$^{-2}$), and ERA5 and ICON underestimate moderate LWP. ICON also overestimates high LWP, resulting in overestimated average LWP.

@@

Fig.~\ref{fig:rs-thetav-hur-cloud} shows $\theta_v$ and RH profiles for profiles containing fog, cloud at 500 m, and cloud at 1.5 km. These situations are characterized by particular cloud biases as identified in the lidar cloud occurrence analysis. The rationale is to examine $\theta_v$ and RH associated with these situations. Foggy situations are characterized by a rapid increase of $\theta_v$ with height and an observed average RH of about 90\% near the surface (Fig.~\ref{fig:rs-thetav-hur-cloud}a). In contrast, the models simulate higher RH in the first 100~m under foggy conditions by several percentage points. In situations with clouds occurring at 500 m, $\theta_v$ is relatively constant between the surface and 500 m (Fig.~\ref{fig:rs-thetav-hur-cloud}b), as expected for convectively driven clouds. The observed RH peaks at 500 m at about 90\%. The models, however, simulated higher RH between the surface and 500 m under these conditions. ICON and ERA5 show a stronger decrease of RH above this height than observations, and ERA5 shows more strongly stable stratification. Unlike the foggy and 500-m cloud situations, situations with clouds at 1.5 km do not have a flat $\theta_v$ with height. This indicates that unlike the former, clouds at 1.5 km are not (or not as strongly) convectively driven. As expected, RH in these situations peaks at 1.5 km at about 85\% in observations. In the models, this peak is much less pronounced.

@@

Fig.~\ref{fig:rs-thetav-hist} shows histograms the same as the previous figure, but for $\theta_v$. They show a more complex picture, characterized by a central peak at about 0°C near the surface, increasing to about 5°C at 2 km (Fig.~\ref{fig:rs-thetav-hist}a). For cloudy bins, the central peak is generally more constant with height and even shows a minimum in $\theta_v$ at about 500 m (Fig.~\ref{fig:rs-thetav-hist}i). This is indicative of convection being associated with clouds at these heights, which results in flat $\theta_v$ profiles. In the reanalyses, in the first 200 m, values slightly above 0°C are associated with more clear bins than in observations, and values slightly below 0°C with fewer (Fig.~\ref{fig:rs-thetav-hist}g--h). Conversely, the opposite is true for cloudy bins (Fig.~\ref{fig:rs-thetav-hist}k--l). Situations with 0°C near-surface air temperature might occur predominantly when open ocean surface keeps the near-surface air temperature close to 0°C under otherwise colder air mass conditions, such as under cold advection. ICON displays a notable bias above about 1 km, where the central peak is strongly underestimated (Fig.~\ref{fig:rs-thetav-hist}j). Instead, these heights and values of $\theta_v$ are more associated with clear bins (Fig.~\ref{fig:rs-thetav-hist}f). This might be related to the strong underestimation of cloud occurrence at these heights.

@@

Let us consider the main limitations of the presented results. The spatial coverage of our dataset does not include most parts of the Indian Ocean and Pacific Ocean sectors of the SO. Even though climatological features of the SO are typically relatively uniform zonally, variations exist, such as those related to the Antarctic Peninsula and the southern tip of South America. The voyages were mostly undertaken in the Austral summer months and only rarely in the winter months, due to the poor accessibility of this region during winter. Therefore, our results are likely representative of summer and, to a lesser extent, spring and autumn conditions. Ship access to sea ice covered areas of the SO is also limited. Cloud regimes and phase in the region are seasonally variable \cite{danker2022}.

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations such as the ENSO in the observations and ICON. The interannual variability in cloud occurrence in ICON can be seen in Fig.~\ref{fig:cloud-occurrence-panel}, where each year in ICON is represented by a separate line. As could be expected, the interannual variability tends to be substantially smaller than the biases and thus is unlikely to have a strong impact on the main findings.

@@

Supercooled liquid clouds (liquid clouds under subzero temperature) commonly occur over the SO. In our analysis of LWP and IWP, we see that both phases are abundant. Because liquid water droplets are typically smaller and more numerous than ice crystals in cold clouds, they attenuate a greater amount of the lidar radiation. Clouds with a relatively modest optical thickness of 1.7 can attenuate the lidar signal for a detection at 2 km using an instrument with noise properties like the Vaisala CL31 (Section 2.4). While supercooled liquid clouds and their attenuation are accounted for by the lidar simulator, they can strongly attenuate the signal and cause artificially low values of cloud occurrence at higher altitudes. For example, we found that cloud occurrence at 1.5 km is underestimated in ICON and underlying clouds are overestimated. However, this can also mean that clouds at 1.5 km are present in the model but the signal is too attenuated by the lower clouds in the model, but not in the observations, where the underlying clouds are not as pronounced.

>>>>

\begin{keypoints}
\item Large cloud biases are present in the global storm-resolving model but are lower in several key aspects than in the reanalyses.
\item The models tend to underestimate total cloud fraction and very low-level ($<$~250~m) cloud and fog, while overestimating cloud occurrence peak at 500 m.
\item A ``too few, too bright'' problem of underestimated cloud fraction, compensated by overestimated cloud albedo, is present in MERRA-2.
\end{keypoints}

@@

Apart from lidar observations, radiosondes were launched on weather balloons at regular synoptic times on the RV \emph{Polarstern}, MARCUS, NBP17024, TAN1702, and TAN1802 campaigns, measuring pressure, temperature, relative humidity (RH), and the global navigation satellite system coordinates. In total, about 2300 radiosonde profiles south of 40°S were available. Spatially and temporally collocated profiles were taken from the models. Because the time period covered by the ICON model output (2021--2024) was different from the time period covered by the observations (2010--2021), when comparing with ICON, we first had to remap the observation time to model time by taking the same time relative to the start of the year. Consequently, we also had four virtual/model profiles (one for each year from 2021 to 2024) for each observed profile. Derived thermodynamic [virtual potential temperature ($\theta_v$), lifting condensation level (LCL), etc.] and dynamic physical quantities (wind speed and direction) for the measured vertical profiles were calculated with the program radiosonde tool [rstool; \citeA{rstool}]. Surface meteorological quantities were measured continuously by an onboard automatic weather station or individual instruments.

@@

An overview of the analyzed campaigns (voyages and stations). Start, end, and the number of days (UTC; inclusive) refer to the time period when the vessel was south of 40°S. Abbreviations: ceilometer (ceil.), Australia (AU), New Zealand (NZ), South America (SA), Atlantic Ocean (AO), and Africa (AF). The number of days is rounded to the nearest integer. CL51/31 indicates CL51 configured to emulate CL31. Missing days in the ceilometer data were HMNZSW16 (7 days): 24--27 November, 10~December, and 16--17~December 2016; MARCUS (3 days): 8, 10~November, and 10~December 2017; MICRE (9~days): 7--8, 29~June, 5, 16~July, 15~August, 17~October 2016, 11~February, and 21~March 2017; and TAN1502 (1~day): 24~January.

@@

The Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2) is a reanalysis produced by the Global Modeling and Assimilation Office at the NASA Goddard Space Flight Center \cite{gelaro2017}. It uses version 5.12.4 of the Goddard Earth Observing System (GEOS) atmospheric model \cite{rienecker2008,molod2015}. Non-convective clouds (condensation, autoconversion, and evaporation) are parameterized using a prognostic scheme \cite{bacmeister2006}, and sub-grid cloud fraction is determined using total water distribution and a critical RH threshold. The reanalysis output analyzed here is available at a spatial resolution of 0.5° of latitude and 0.625° of longitude, which is about 56 km in the north--south direction and 35 km in the east--west direction at 60°S. The number of vertical model levels is 72. Here, we use the following products: 1-hourly instantaneous 2D single-level diagnostics (M2I1NXASM) for 2-m temperature and humidity; 3-hourly instantaneous 3D assimilated meteorological fields (M2I3NVASM) for cloud quantities, pressure, and temperature; 1-hourly average 2D surface flux diagnostics (M2T1NXFLX) for precipitation; and 1-hourly average 2D radiation diagnostics (M2T1NXRAD) for radiation quantities \cite{merra2}. Vertically resolved fields in M2I3NVASM start at a height of about 60 m, which limits our analysis of fog and very low-level ($<$~250~m) clouds in this reanalysis.

@@

ERA5 \cite{era5} is a reanalysis produced by the ECMWF. It is based on a numerical weather prediction model IFS version CY41R2. It uses the \citeA{tiedtke1993} prognostic cloud scheme and \citeA{forbes2014} for mixed-phase clouds. The horizontal resolution is 0.25° in latitude and longitude, which is about 28~km in the north--south direction and 14~km in the east--west direction at 60°S. Internally, the model uses 137 vertical levels. Here, we use output at 1-hourly instantaneous time intervals, except for radiation quantities, which are accumulations (from these we calculate daily means). Vertically resolved quantities are made available on 37 pressure levels.

@@

In our analysis, we partitioned our dataset by cyclonic activity and stability into multiple subsets to evaluate cloud biases in the context of the main physical controlling processes. The SO is a region of the occurrence of both extratropical and polar cyclones. Cyclonic activity results in cloud formation at the air mass boundaries along the cold and warm fronts, as well as inside the cold sector, after a passing cold sector destabilizes the atmosphere relative to the surface temperature. In the cold front and cold sector, clouds are convectively driven, including deep convection, and the advection of colder air masses over warmer ocean surfaces can trigger convection and subsequent cloud formation. In contrast, warm advection can trigger fog or cloud formation by boundary layer air cooled by the ocean surface until it reaches saturation. More quiescent areas outside of cyclones can also be associated with clouds. These can be, for example, associated with clouds formed by warm or cold advection outside of cyclones, persistent clouds, clouds formed due to diurnal heating or cooling, or clouds formed due to ocean currents. Boundary layer stability can be expected to be associated with clouds by either allowing convection and turbulence under weak stability, inhibiting convection turbulence under strong stability, and by capping inversion controlling the cloud top height or trapping moist air near the surface and preventing fog dispersion. Therefore, dividing our dataset by these subsets allows us to quantify model biases associated with some of the main physical processes controlling cloud formation, persistence, and dissipation. Other methods of subsetting, such as using the ISCCP pressure--optical thickness diagram \cite{rossow1991,rossow1999,hahn2001} to separate profiles by cloud regimes and other cloud regime classifications \cite{oreopoulos2016,schuddeboom2018}, would be feasible.

We partitioned our data into two mutually exclusive subsets by cyclonic activity. For this purpose, we used a cyclone tracking algorithm to identify extratropical and polar cyclones (ECs and PCs) over the SO in the reanalysis and ICON data. We used the open-source cyclone tracking package CyTRACK \cite{perez-alarcon2024}. Generally, what constitutes an EC is considered relatively arbitrary due to the very large variability of ECs \cite{neu2013}. The CyTRACK algorithm uses mean sea level pressure and wind speed thresholds as well as tracking across time steps to identify cyclone centers and their radii in each time step. With this information, we could classify every location at a given time as either cyclonic or non-cyclonic. Due to a relatively small total area covered by cyclones, as identified by the cyclone center and radius, for every time step and cyclone, we defined a cyclonic area as a circle of double the radius identified by CyTRACK centered at the cyclone center. All other areas were defined as non-cyclonic. For identifying cyclones in the observations and the reanalyses, ERA5 pressure and wind fields were used as the input to CyTRACK. This is justified by the fact that the large-scale pressure and wind fields in ERA5 are likely sufficiently close to reality. \citeA{mcerlich2023} have shown that wind is simulated well in ERA5 relative to the WindSat polarimetric microwave radiometer measurements \cite{meissner2009}. For identifying cyclones in ICON, its own pressure and wind fields were used as the input to CyTRACK because ICON is free-running, and thus the pressure and wind fields are different from reality. Subsetting by proximity to cyclones is a relatively crude measure because it does not take into account the different sectors of cyclones, which are commonly associated with different weather situations. However, this was a choice made for simplicity of the analysis, given the quantity of data. \citeA{konstali2024} performed a more complex attribution of precipitation to individual cyclone features.

@@

Fig.~\ref{fig:cyclone-stability}a and b show the geographical distribution of the fraction of cyclonic days as determined by the cyclone tracking algorithm applied to the ERA5 reanalysis and ICON data (Section~\ref{sec:cyclone-stability}). As expected, the strongest cyclonic activity is in the high-latitude SO zone and is relatively zonally symmetric at all latitudes. The pattern matches reasonably well with \citeA{hoskins2005}. While both reanalysis and ICON agree within about 8\% in most areas, ICON is prevailingly more cyclonic by about 4\%. There are clear differences, particularly in the highest occurrence rate regions, such as around Cape Adare, which is up to 20\% more cyclonic in ICON, and the Weddell and Bellingshausen Seas, where ICON is less cyclonic by up to 10\%. These differences might, however, stem from the relatively short time periods of comparison (4 years) and the fact that ICON is free-running.

@@

When the data are subset by latitude (Fig.~\ref{fig:cloud-occurrence}b, c), we see that the low-latitude SO zone (40--55°S) displays a stronger peak of cloud occurrence near the surface than the high-latitude SO zone (between 55°S and the Antarctic coast), and this could be because higher latitudes have a greater prevalence of weakly stable profiles (Fig.~\ref{fig:cyclone-stability}c, d), although more stable profiles populate regions south of 65°S close to the Antarctic coast. Cyclonic activity is also stronger in high-latitude SO, which is typically associated with shallow or deep convection rather than the very stable stratification necessary for fog formation. The low- and high-latitude SO zones show similar biases in models as in the general case, but ERA5 does not overestimate the peak in the low-latitude SO zone (very low-level cloud occurrence is still strongly underestimated).

When the data are subset as either cyclonic or non-cyclonic situations (Fig.~\ref{fig:cloud-occurrence}d, e), we see that the cyclonic situations have a larger amount of observed cloudiness, including peak and total cloud fraction, by about 10\%. In the cyclonic situations, the model vertical profiles of cloud occurrence compare well with observations, but they peak higher by about 200 m and are larger by about 8\%. The reanalyses tend to underestimate cloud occurrence above 1 km by about 5\% and near the surface by about 15\%. Non-cyclonic situations are similar to the general case, also because they form the majority of analyzed profiles (83\%).

When the data are subset by stability (Fig.~\ref{fig:cloud-occurrence}f, g), as defined in Section~\ref{sec:cyclone-stability}, we see that in situations of strong stability, cloud occurrence peaks strongly near the surface in observations, compared to situations of weak stability, where the peak is more diffuse between 0 and 1~km. Physically, conditions of strong stability are associated with the formation of advection fog, such as in situations of warm air advection from the north over a colder sea surface, thus inducing fog formation by cooling of the warm and humid air by the cold surface. In situations of strong stability, the models have smaller biases than in weak stability, with an overestimated peak of up to 12\%, underestimated cloud occurrence above 1~km by up to 5\%, and underestimated cloud occurrence near the surface by about 10\% in the reanalyses but not ICON. In situations of weak stability, the bias in ICON is very pronounced, with a much larger peak in cloud occurrence at about 500~m; the reanalyses underestimate cloud occurrence below 1~km, especially near the surface; and MERRA-2 underestimates cloud occurrence more strongly at almost all heights.

@@

\begin{figure}[p!]
\centering
\includegraphics[width=\textwidth]{img/cl_agg_rev1.pdf}
\caption{
Cloud occurrence by height calculated as the average of all voyages and stations and lifting condensation level (LCL) distribution. The LCL is derived from radiosonde profiles and equivalent model profiles, which were not available for all voyages and times. The total cloud fraction (CF), average outgoing shortwave (SW) and longwave (LW) radiation, and the relative frequency of occurrence (RFO) are shown. The bands are the 16$^\mathrm{th}$--84$^\mathrm{th}$ percentile, calculated from the set of all voyages and stations.
}
\label{fig:cloud-occurrence}
\end{figure}

@@

We also analyzed the daily cloud cover (total cloud fraction) distribution. This is a measure of cloudiness, irrespective of height, calculated over the course of a day (UTC). A cloud detected at any height means that the lidar profile was classified as cloudy; otherwise, it was classified as a clear sky. When all profiles in a day are taken together, the cloud cover for the day is defined as the fraction of cloudy profiles in the total number of profiles. It is expressed in oktas (multiples of 1/8), reflecting the 3-hourly model output of MERRA-2 and ICON, i.e., 8 times per day. The same calculation is done for the lidar observations as for the simulated lidar profiles. We use the term ``okta'' independently of its use in instantaneous synoptic observations, and here it simply means 1/8 (0.125) of the daily cloud cover.

@@

Fig.~\ref{fig:stats-hist-si} shows liquid water path (LWP) and ice water path (IWP) distributions as histograms and their corresponding averages. LWP and IWP are calculated from the mass of water in the column divided by the area of the column, i.e., not just the area of the cloudy portion of the column, as in some definitions. The available observational satellite reference for LWP and IWP over high latitudes is unfortunately very uncertain due to a high solar zenith angle and the inability of passive visible and infrared retrievals to detect phase below the cloud top of mixed-phase clouds \cite{huang2006,greenwald2009,seethala2010,eliasson2011,duncan2018,khanal2020}, and this limits our comparison. The LWP distribution shows that all models overestimate cases with near-zero LWP (Fig.~\ref{fig:stats-hist-si}a), which relates to the underestimated total cloud fraction. MERRA-2 shows quite overestimated high-LWP situations, which is most likely related to the too few, too bright problem of simulating lower total cloud fraction but clouds with higher LWP to compensate. IWP (Fig.~\ref{fig:stats-hist-si}b) is somewhat less important radiatively than LWP because of the typically larger and less numerous hydrometeors. Similarly to LWP, the models overestimate situations with near-zero IWP. ERA5 is otherwise simulating the IWP distribution well, but ICON and MERRA-2 underestimate IWP. In the cloudy situations (Fig.~\ref{fig:stats-hist-si}c, d), it can be seen more distinctly that MERRA-2 overestimates moderate (0.05--0.15 kg m$^{-2}$) and high LWP (over 0.15 kg m$^{-2}$), and ERA5 and ICON underestimate moderate LWP. ICON also overestimates high LWP, resulting in overestimated average LWP.

@@

Fig.~\ref{fig:rs-thetav-hur-cloud} shows $\theta_v$ and RH profiles for profiles containing fog, cloud at 500 m, and cloud at 1.5 km. These situations are characterized by particular cloud biases as identified in the lidar cloud occurrence analysis. The rationale is to examine $\theta_v$ and RH associated with these situations. Foggy situations are characterized by a rapid increase of $\theta_v$ with height and an observed average RH of about 90\% near the surface (Fig.~\ref{fig:rs-thetav-hur-cloud}a). In contrast, the models simulate higher RH in the first 100~m under foggy conditions by several percentage points. In situations with clouds occurring at 500 m, $\theta_v$ is relatively constant between the surface and 500 m (Fig.~\ref{fig:rs-thetav-hur-cloud}b), as expected for convectively driven clouds. The observed RH peaks at 500 m at about 90\%. The models, however, simulated higher RH between the surface and 500 m under these conditions. ICON and ERA5 show a stronger decrease of RH above this height than observations, and ERA5 shows more strongly stable stratification. Unlike the foggy and 500-m cloud situations, situations with clouds at 1.5 km do not have a flat $\theta_v$ with height. This indicates that, unlike the former, clouds at 1.5 km are not (or not as strongly) convectively driven. As expected, RH in these situations peaks at 1.5 km at about 85\% in observations. In the models, this peak is much less pronounced.

@@

Fig.~\ref{fig:rs-thetav-hist} shows histograms the same as the previous figure, but for $\theta_v$. They show a more complex picture, characterized by a central peak at about 0°C near the surface, increasing to about 5°C at 2 km (Fig.~\ref{fig:rs-thetav-hist}a). For cloudy bins, the central peak is generally more constant with height and even shows a minimum in $\theta_v$ at about 500 m (Fig.~\ref{fig:rs-thetav-hist}i). This is indicative of convection being associated with clouds at these heights, which results in flat $\theta_v$ profiles. In the reanalyses, in the first 200 m, values slightly above 0°C are associated with more clear bins than in observations, and values slightly below 0°C with fewer (Fig.~\ref{fig:rs-thetav-hist}g--h). Conversely, the opposite is true for cloudy bins (Fig.~\ref{fig:rs-thetav-hist}k--l). Situations with 0°C near-surface air temperature might occur predominantly when an open ocean surface keeps the near-surface air temperature close to 0°C under otherwise colder air mass conditions, such as under cold advection. ICON displays a notable bias above about 1 km, where the central peak is strongly underestimated (Fig.~\ref{fig:rs-thetav-hist}j). Instead, these heights and values of $\theta_v$ are more associated with clear bins (Fig.~\ref{fig:rs-thetav-hist}f). This might be related to the strong underestimation of cloud occurrence at these heights.

@@

Let us consider the main limitations of the presented results. The spatial coverage of our dataset does not include most parts of the Indian Ocean and Pacific Ocean sectors of the SO. Even though climatological features of the SO are typically relatively uniform zonally, variations exist, such as those related to the Antarctic Peninsula and the southern tip of South America. The voyages were mostly undertaken in the Austral summer months and only rarely in the winter months, due to the poor accessibility of this region during winter. Therefore, our results are likely representative of summer and, to a lesser extent, spring and autumn conditions. Ship access to sea-ice-covered areas of the SO is also limited. Cloud regimes and phases in the region are seasonally variable \cite{danker2022}.

The time period of ICON is relatively short, with only four full years of simulation available. Moreover, the simulation is free-running and ocean-coupled, which means that observations had to be temporally mapped to this time period (at the same time relative to the start of the year) for the comparison. For these reasons, one can expect the results to be slightly different due to reasons unrelated to model biases, such as different weather conditions, partially accounted for by the cyclone and stability subsetting, and the phase of climate oscillations, such as the ENSO in the observations and ICON. The interannual variability in cloud occurrence in ICON can be seen in Fig.~\ref{fig:cloud-occurrence-panel}, where each year in ICON is represented by a separate line. As could be expected, the interannual variability tends to be substantially smaller than the biases and thus is unlikely to have a strong impact on the main findings.

@@

Supercooled liquid clouds (liquid clouds under subzero temperature) commonly occur over the SO. In our analysis of LWP and IWP, we see that both phases are abundant. Because liquid water droplets are typically smaller and more numerous than ice crystals in cold clouds, they attenuate a greater amount of the lidar radiation. Clouds with a relatively modest optical thickness of 1.7 can attenuate the lidar signal for a detection at 2 km using an instrument with noise properties like the Vaisala CL31 (Section 2.4). While supercooled liquid clouds and their attenuation are accounted for by the lidar simulator, they can strongly attenuate the signal and cause artificially low values of cloud occurrence at higher altitudes. For example, we found that cloud occurrence at 1.5 km is underestimated in ICON and underlying clouds are overestimated. However, this can also mean that clouds at 1.5 km are present in the model, but the signal is too attenuated by the lower clouds in the model, but not in the observations, where the underlying clouds are not as pronounced.
